---
title: "Explore US Bikeshare Data"
author: "Kirill Shish"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(ggplot2)
library(dplyr)
library(lubridate)

## Introduction

This project analyzes bike rental data for the first half of 2017 for three US cities: **Chicago**, **New York** and **Washington**.

Objectives of the analysis:
- Determine the most popular travel time.
- Identify the busiest stations and routes.
- Compare the duration of trips between cities.
- Study the demographic characteristics of users.

---

## Data preparation

```{r}
# Data loading and preprocessing function
load_and_clean <- function(city) {
  filename <- paste0(city, ".csv")
  df <- read.csv(filename)
  df$Start.Time <- ymd_hms(df$Start.Time)
  df$month <- month(df$Start.Time, label = TRUE, abbr = FALSE)
  df$day_of_week <- wday(df$Start.Time, label = TRUE, abbr = FALSE)
  df$hour <- hour(df$Start.Time)
  df$route <- paste(df$Start.Station, "->", df$End.Station)
  df$city <- city
  return(df)
}

# Loading all three cities
chi <- load_and_clean("chicago")
nyc <- load_and_clean("new-york-city")
was <- load_and_clean("washington")

# Combining them for a general analysis
all_cities <- bind_rows(
  chi %>% mutate(City = "Chicago"),
  nyc %>% mutate(City = "New York"),
  was %>% mutate(City = "Washington")
)
all_cities$duration_min <- all_cities$Trip.Duration / 60
```

The data has been uploaded successfully. Total trips: **`r nrow(all_cities)`**.

---

## 1. Popular travel times

```{r}
cat("=== ЧИКАГО ===\n")
cat("The most frequent month:       ", names(sort(table(chi$month), decreasing = TRUE)[1]), "\n")
cat("The most frequent day of the week: ", names(sort(table(chi$day_of_week), decreasing = TRUE)[1]), "\n")
cat("The most frequent hour:          ", names(sort(table(chi$hour), decreasing = TRUE)[1]), ":00\n\n")

cat("=== New York ===\n")
cat("The most frequent month:       ", names(sort(table(nyc$month), decreasing = TRUE)[1]), "\n")
cat("The most frequent day of the week: ", names(sort(table(nyc$day_of_week), decreasing = TRUE)[1]), "\n")
cat("The most frequent hour:          ", names(sort(table(nyc$hour), decreasing = TRUE)[1]), ":00\n\n")

cat("===Washington ===\n")
cat("The most frequent month:       ", names(sort(table(was$month), decreasing = TRUE)[1]), "\n")
cat("The most frequent day of the week: ", names(sort(table(was$day_of_week), decreasing = TRUE)[1]), "\n")
cat("The most frequent hour:          ", names(sort(table(was$hour), decreasing = TRUE)[1]), ":00\n")
```

### 📊 График: Distribution of trips by hours of the day (New York)
```{r}
ggplot(nyc, aes(x = hour)) +
  geom_bar(fill = "#2E86AB", color = "white", alpha = 0.8) +
  labs(title = "Hourly Distribution of Trips (New York)",
       subtitle = "Data for the first half of 2017",
       x = "Hour of Day (0 = midnight, 23 = 11 PM)",
       y = "Number of Trips") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  theme_minimal(base_size = 12)
```

**Output:** Peak travel occurs in the morning (7-9) and evening (16-18) hours — this corresponds to rush hours when people travel to and from work.
---

## 2. Popular stations and routes

```{r}
cat("=== CHICAGO ===\n")
cat("Most frequent start station:\n  ", names(sort(table(chi$Start.Station), decreasing = TRUE)[1]), "\n")
cat("Most frequent route:\n  ", names(sort(table(chi$route), decreasing = TRUE)[1]), "\n\n")

cat("=== NEW YORK ===\n")
cat("Most frequent start station:\n  ", names(sort(table(nyc$Start.Station), decreasing = TRUE)[1]), "\n")
cat("Most frequent route:\n  ", names(sort(table(nyc$route), decreasing = TRUE)[1]), "\n\n")

cat("=== WASHINGTON ===\n")
cat("Most frequent start station:\n  ", names(sort(table(was$Start.Station), decreasing = TRUE)[1]), "\n")
cat("Most frequent route:\n  ", names(sort(table(was$route), decreasing = TRUE)[1]), "\n")
```

**Output:** The most popular stations are located in the central and tourist areas of the cities.

---

## 3. Duration of trips

```{r}
cat("=== AVERAGE DURATION (minutes) ===\n")
cat("Chicago:     ", round(mean(chi$Trip.Duration / 60), 2), "\n")
cat("New York:   ", round(mean(nyc$Trip.Duration / 60), 2), "\n")
cat("Washington:  ", round(mean(was$Trip.Duration / 60), 2), "\n\n")

cat("=== TOTAL TIME (hours) ===\n")
cat("Chicago:     ", round(sum(chi$Trip.Duration) / 3600, 1), "\n")
cat("New York:   ", round(sum(nyc$Trip.Duration) / 3600, 1), "\n")
cat("Washington:  ", round(sum(was$Trip.Duration) / 3600, 1), "\n")
```

### 📊 Graph: Comparison of the duration of trips by city
```{r}
ggplot(all_cities, aes(x = City, y = duration_min, fill = City)) +
  geom_boxplot(alpha = 0.7, outlier.color = "gray60", outlier.size = 0.5) +
  labs(title = "Comparison of Trip Durations by City",
       x = "City",
       y = "Trip Duration (minutes)") +
  scale_fill_manual(values = c("Chicago" = "#C73E1D", 
                               "New York" = "#2E86AB", 
                               "Washington" = "#6A994E")) +
  coord_cartesian(ylim = c(0, 30)) +  # Limit Y axis for visual clarity (majority of trips are under 30 mins)
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

**Output:** The average trip duration is 10-15 minutes. Trips in Washington are slightly longer than in Chicago and New York.
---

## 4.User Information

```{r}
cat("=== USER TYPES ===\n")
table(all_cities$City, all_cities$User.Type)

cat("\n=== USER GENDER (Chicago and New York) ===\n")
cat("Chicago:\n"); print(table(chi$Gender))
cat("\nNew York:\n"); print(table(nyc$Gender))
cat("\n(Gender and birth year data are missing for Washington)\n")

cat("\n=== BIRTH YEAR (Chicago and New York) ===\n")
cat("Chicago:\n")
cat("  Earliest: ", min(chi$Birth.Year, na.rm = TRUE), "\n")
cat("  Latest:   ", max(chi$Birth.Year, na.rm = TRUE), "\n")
cat("  Most frequent: ", names(sort(table(chi$Birth.Year), decreasing = TRUE)[1]), "\n\n")
cat("New York:\n")
cat("  Earliest: ", min(nyc$Birth.Year, na.rm = TRUE), "\n")
cat("  Latest:   ", max(nyc$Birth.Year, na.rm = TRUE), "\n")
cat("  Most frequent: ", names(sort(table(nyc$Birth.Year), decreasing = TRUE)[1]), "\n")
```

### 📊 Graph: Types of users by city
```{r}
ggplot(all_cities, aes(x = City, fill = User.Type)) +
  geom_bar(position = "dodge", color = "white") +
  labs(title = "Comparison of User Types by City",
       x = "City",
       y = "Number of Trips",
       fill = "User Type") +
  scale_fill_manual(values = c("Subscriber" = "#F18F01", "Customer" = "#006E90")) +
  theme_minimal(base_size = 12)
```

**Output:** The vast majority of users are subscribers. Men predominate among those who indicated their gender. The most frequent year of birth is the end of the 1980s.

---

## Conclusion

During the analysis of bicycle rental data for 2017, the following patterns were identified:

1. **Time:** The greatest activity is observed during the morning and evening rush hours, as well as during the warmer months (June-August).
2. **Stations:** The most popular stations are located in the business and tourist centers of the cities.
3. **Duration:** Most trips are short (up to 15 minutes), which is typical for urban transport.
4. **Users:** The base consists of male subscribers born in the 1980s.

This analysis can be useful for optimizing the distribution of bicycles across stations and planning marketing campaigns.
